---
title: "Final Project Presentation"
author: "Ellen Liu"
date: "12-11-2020"
output: html_notebook
---

# Guiding Question: How gun-related violence affect the United States?
  As the U.S. is one of the few countries in which the right to bear arms is constitutionally protected. A consequence of this is that the level of gun voilence is very high compared to other countries. So for this project, I want to performed some deep explorations on this Dataset of gun violence incidents provided by Gun Violence Archive (GVA). I expect this analysis help us have a more comprhensive understanding about the gun violence in U.S. and take gun control more seriously.
  
## Monthly incidents for each year
```{r}
library(scales)
quarterly<-
Gun_violence_new %>%
  filter(year!= 2013)  %>%
  group_by(year,month) %>%
  summarise(monthly= n()) 
quarterly
```
  
  
```{r}
quarterly%>%
ggplot(aes(x=month, y=monthly)) + geom_bar(stat='identity')  + facet_wrap(.~year,ncol=2) + labs(x='monthly', y='Number of incidents')+scale_x_continuous(breaks = pretty_breaks())+theme_linedraw() 
```


## Incidents relative to the State population size

```{r}
#Incidents of each State in descending order
topState <- 
  Gun_violence %>%
  group_by(state) %>%
  summarise(case_count= n()) %>%
   arrange(desc(case_count))

#Incidents relative to the State population size
crimedensity<-
  topState %>%
left_join(useful_ZipGeography, by = c("state"= "State")) %>%
  mutate(per_million=case_count/(population/1000000))


crimedensity
```


```{r}
crimedensity %>%
  ggplot(aes(y=reorder(state, per_million), x=per_million, fill=per_million, text=state,width=.7))+geom_bar(stat='identity')+theme(legend.position="none")+labs(x='Incidents per million people', y='State')+theme_linedraw() 
```

